---
title: "La Ronge Tree Heights"
author: "SMurphy"
date: "2023-10-15"
output: 
  github_document:
    toc: TRUE
    toc_depth: 5
    number_sections: FALSE
    df_print: tibble
---

```{r setup, echo=FALSE, message=FALSE,warning=FALSE, error=FALSE}
library(sf)
library(RColorBrewer)
library(lidR)
library(ForestTools)
library(rgl)
library(pandocfilters)
library(rmarkdown)
library(formatR)
library(gitignore)
library(tinytex)
library(knitr)
library(raster)
library(webdriver)
library(webshot)
library(webshot2)
library(terra)
library(matlab)
#webshot::install_phantomjs(force = TRUE)
knit_hooks$set(webgl = hook_webgl)
knit_hooks$set(rgl.static = hook_rgl)
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, error=FALSE, message = FALSE)
set.seed(23)
```

## Data Source

A digital surface model was acquired in raster form with 1m resolution from the HRDEM repository for La Ronge surrounding area in the following location: https://open.canada.ca/data/en/dataset/957782bf-847c-4644-a757-e383c0057995/resource/83300867-10b0-4c9d-a2b6-3921f2a07dcb#additional-info

```{r, rgl.static=TRUE, cache=TRUE, eval=TRUE, echo=TRUE}
dsm = raster::raster("/media/seamus/USB1/la_ronge/dsm_la_ronge.tif", select = 'xyzcr', filter = '-drop_class 19')
plot(dsm)
```

## Tree top detection 

We adopt below Popescu and Wynne's function (2004) which was developed in pine and spruce forests to define a variable window filter algorithm. We assume that dominant, co-dominant, intermediate, and suppressed trees are of varying sizes and of varying crown widths. That is, taller trees, have wider crowns. Therefore, we apply a simple linear function (wf_plowright) to adapt the radius of the search window according to raster cell height. With more resources available, a local radius to height curve can be derived from the site's dominant allometry. 

To exclude low-lying underbrush or other spurious treetops, the variable window function is also set with a minimum height of 2 m using the minHeight argument as follows. 

```{r, echo=TRUE, eval=TRUE}
kernel <- matrix(1,3,3)
wf_Popescu<-function(x){ 
  a=0.05
  b=0.6 
  y<-a*x+b 
  return(y)}
heights <- seq(0,40,0.5)
window_Popescu <- wf_Popescu(heights)
plot(heights, window_Popescu, type = "l", ylim = c(0,12), xlab="point elevation (m)", ylab="window diameter (m)", main='Variable Window Function (Popescu & Wynne 2004)')
```

To improve performance of the above function, we apply a smoothing function to our DSM. With these treated cells, we then apply the upside-down watershed algorithm. 

```{r, echo=TRUE, eval=TRUE}
dsm_smooth = focal(dsm, w = kernel, fun = median, na.rm = TRUE) 
chm = ForestTools::vwf(CHM = dsm_smooth, winFun = wf_plowright, minHeight = 2)
chm_sp = as_Spatial(chm)
chm_sf = st_as_sf(chm_sp)
writeOGR(chm_sp, "/media/seamus/USB1/la_ronge", "chm", driver = "ESRI Shapefile") 
```