---
title: "La Ronge Tree Heights"
author: "SMurphy"
date: "2023-10-15"
output: 
  html_document:
    toc: TRUE
    toc_depth: 5
    number_sections: FALSE
    df_print: tibble
---

```{r setup, echo=FALSE, message=FALSE,warning=FALSE, error=FALSE}
library(sf)
library(ggplot2)
library(tidyterra)
library(RColorBrewer)
library(lidR)
library(ForestTools)
library(rgl)
library(pandocfilters)
library(rmarkdown)
library(formatR)
library(gitignore)
library(tinytex)
library(knitr)
library(raster)
library(webdriver)
library(webshot)
library(webshot2)
library(terra)
library(matlab)
library(spatstat.geom)
library(magic)
library(geometry)
library(deldir)
library(bitops)
library(rLiDAR)
#webshot::install_phantomjs(force = TRUE)
knit_hooks$set(webgl = hook_webgl)
knit_hooks$set(rgl.static = hook_rgl)
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, error=FALSE, message = FALSE)
set.seed(23)
```

## Intro: Tree height variability

We apply the “Height Variation Hypothesis” and associated methods to estimate tree height heterogeneity (MacArthur and MacArthu, 1961). Due to varied factors regarding yield class, soil moisture, browsing, species diversity, stocking density, line of sight and clinometer errors, tree height variability presents a key challenge to upholding accuracy targets in forest inventory operations. The following tool aims to assist inventory crews to identify areas of high tree height heterogeneity, where there may be need for an increase in number of sample plots or for a decrease in the size of sampling units. Using LiDAR derived tree metrics, we classify forest areas according simple standard deviation values of tree height to produce maps of Height Heterogeneity Areas, or HHA's.


## Data source

A digital surface model and digital elevation model were acquired in raster form and with 1m resolution from the HRDEM repository for the La Ronge surrounding area, as shown in the following location: https://open.canada.ca/data/en/dataset/957782bf-847c-4644-a757-e383c0057995/resource/83300867-10b0-4c9d-a2b6-3921f2a07dcb#additional-info

```{r, rgl.static=TRUE, fig.show='hold', out.width="33%", cache=TRUE, eval=TRUE, echo=TRUE}
dsm = raster::raster("/media/seamus/USB1/la_ronge/dsm_la_ronge.tif", select = 'xyzcr', filter = '-drop_class 19')
dem = raster::raster("/media/seamus/USB1/la_ronge/dem_la_ronge.tif", select = 'xyzcr', filter = '-drop_class 19')
dsm_sr = terra::rast(dsm)
dem_sr = terra::rast(dem)
chm_sr = dsm_sr - dem_sr
chm = raster::raster(chm_sr)
plot(dsm_sr, main="DSM (metres a.s.l.)", cex.main=0.8)
plot(dem_sr, main="DEM (metres a.s.l.)", cex.main=0.8)
plot(chm_sr, main="CHM (unprocessed)", cex.main=0.8)
```

## Tree top detection 

We adopt below Popescu and Wynne's function (2004) which was developed in pine forests to define a variable window filter for detecting tree tops. This is to account for stand variance across dominant, co-dominant, intermediate, and suppressed trees, which present varying heights and varying crown widths. We assume taller trees have wider crowns. Therefore, we apply a simple linear function (wf_Popescu) to adapt the radius of the search window according to canopy height. With more resources available, a local radius to height curve can be derived from the site's dominant allometry. 

To exclude low-lying underbrush or other spurious treetops, the variable window function is also set with a minimum height of 2 m using the minHeight argument as follows. 

```{r, echo=TRUE, eval=TRUE}
kernel <- matrix(1,3,3)
wf_Popescu<-function(x){ 
  a=0.05
  b=0.6 
  y<-a*x+b 
  return(y)}
heights <- seq(0,40,0.5)
window_Popescu <- wf_Popescu(heights)
plot(heights, window_Popescu, type = "l", ylim = c(0,12), xlab="point elevation (m)", ylab="window diameter (m)", main='Variable Window Function (Popescu & Wynne 2004)')
```

To improve algorithm performance, we apply a smoothing operation to our DSM raster. With cells softened, we then apply the upside-down watershed algorithm from the ForestTools package, shown as vwf below. Outputs are then converted from simple features to derive 95% height raster and stem point shapefile in ESRI format using a custom function.   

```{r, fig.show='hold', out.width="50%", echo=TRUE, eval=FALSE}
chm_smooth = focal(chm, w = kernel, fun = median, na.rm = TRUE) 
ttops = ForestTools::vwf(CHM = chm_smooth, winFun = wf_Popescu, minHeight = 2)
ttops$height_95 = ttops$height * 0.95 #avoid re-runs
ttops$height_95_sd = (ttops$height_95 - sd(ttops$height_95))/10
sf::st_write(ttops, "/media/seamus/USB1/la_ronge/chm_la_ronge.shp")
chm_ttops = stars::st_rasterize(ttops %>% dplyr::select(height_95, geometry))
chm_ttops_stdev = stars::st_rasterize(ttops %>% dplyr::select(height_95_sd, geometry))
stars::write_stars(chm_ttops, "/media/seamus/USB1/la_ronge/chm_ttops_raster.tif")
stars::write_stars(chm_ttops_stdev, "/media/seamus/USB1/la_ronge/chm_ttops_stdev.tif")
chm_ttops_raster = raster::raster("/media/seamus/USB1/la_ronge/chm_ttops_raster.tif")
chm_ttops_stdev = raster::raster("/media/seamus/USB1/la_ronge/chm_ttops_stdev.tif")
chm_ttops_spatRast = terra::rast(chm_ttops_raster)
chm_stdev_spatRast = terra::rast(chm_ttops_stdev)

plot(chm_ttops_spatRast, 
     col = height.colors(50), 
     main="Stem-derived CHM (95% Height)")

chm_ttops_spatRast[chm_ttops_spatRast == 0] <- NA
chm_ttops_raster[chm_ttops_raster == 0] <- NA
ggplot() +
  geom_spatraster(data = chm_ttops_spatRast) +
  geom_spatraster_contour(data = chm_ttops_spatRast, breaks = seq(5, 30, 5)) +
  scale_fill_whitebox_c() +
  coord_sf(expand = FALSE) +
  labs(fill = "Tree Height (95%)")

ttops %>% as_tibble() %>% print(n=10)
ttops %>% rename(crownWidth = winRads)

ggplot() +
  geom_sf(data = ttops, aes(color = ttops$hght_95_sd), show.legend = "point", size = 0.05, shape = ".") +
    scale_fill_continuous(palette = "lajolla")
  labs(fill = "Tree Height (95%)")
```

```{r, fig.show='hold', out.width="50%", echo=FALSE, eval=TRUE}
chm_ttops_raster = raster::raster("/media/seamus/USB1/la_ronge/chm_ttops_raster.tif")
chm_ttops_stdev = raster::raster("/media/seamus/USB1/la_ronge/chm_ttops_stdev.tif")
chm_ttops_spatRast = terra::rast(chm_ttops_raster)
chm_stdev_spatRast = terra::rast(chm_ttops_stdev)

plot(chm_ttops_spatRast, 
     col = height.colors(50), 
     main="Stem-derived CHM (95% Height)")

chm_ttops_spatRast[chm_ttops_spatRast == 0] <- NA
chm_ttops_raster[chm_ttops_raster == 0] <- NA
ggplot() +
  geom_spatraster(data = chm_ttops_spatRast) +
  geom_spatraster_contour(data = chm_ttops_spatRast, breaks = seq(5, 30, 5)) +
  scale_fill_whitebox_c() +
  coord_sf(expand = FALSE) +
  labs(fill = "Tree Height (95%)")

ttops = sf::st_read("/media/seamus/USB1/la_ronge/chm_la_ronge.shp")
ttops_df = as.data.frame(ttops)
ttops %>% rename(crownWidth = winRads)
ttops %>% as_tibble() %>% print(n=10)
```

```{r, fig.show='hold', out.width="100%", echo=FALSE, eval=TRUE}
ggplot() +
  geom_sf(data = ttops, aes(color = ttops$hght_95_sd), show.legend = "point", size = 0.05, shape = ".") +
    scale_fill_continuous(palette = "lajolla") + 
  labs(fill = "Tree Height (95%)")
```

## Validation

Results are validated using tree-ID and crown segmentation. These steps check for double-counting and overlapping edges. 

```{r, echo=TRUE, eval=FALSE}
crown_segment = itcSegment::itcIMG(chm_ttops_spatRast,epsg=2957)
summary(crown_segment)
plot(crown_segment,axes=T)
```


## Height heterogenity areas (HHAs)

Using the simple structural diversity index developed by MacArthur and MacArthur (1961), we derived structural diversity classes according to height heterogeneity. Targeting height heterogeneity areas (HHA's), classes showing highest standard deviation were highlighted below in blue. 

Further metrics may consider combining horizontal estimates from within and between sampling units, as well as crown width heterogeneity in areas where there higher concentrations of wolf and veteran trees. Here we simply rely on raster cell boundaries against population mean  as a quick fix. 

```{r, echo=TRUE, eval=TRUE}

print(chm_ttops_stdev)

chm_ttops_raster_20m = terra::aggregate(chm_ttops_raster, fact = 20, fun = mean)
chm_ttops_raster_100m = terra::aggregate(chm_ttops_raster_20m, fact = 5, fun = mean)

reclass_df <- c(0, 5, 1,
                5, 10, 2,
                10, 15, 3,
                15, Inf, 4)

reclass_m <- matrix(reclass_df,
                ncol = 3,
                byrow = TRUE)

chm_classified_20m <- reclassify(chm_ttops_raster_20m,
                     reclass_m)


plot(chm_classified_20m,
     col = c("yellow", "red", "green", "blue"))
legend("topleft",
       legend = c("low variability", "slight variability", "moderate variability", "high variability"),
       fill = c("yellow", "red", "green", "blue"),
       border = FALSE,
       bty = "n",
       cex=0.9) # turn off legend border
```


## References

MacArthur, R. H., & MacArthur, J. W. (1961). On bird species diversity. Ecology, 42(3), 594-598.

Popescu, S. C., & Wynne, R. H. (2004). Seeing the trees in the forest. Photogrammetric Engineering & Remote Sensing, 70(5), 589-604.

Torresani, M., Rocchini, D., Alberti, A., Moudrý, V., Heym, M., Thouverai, E., & Tomelleri, E. (2023). LiDAR GEDI derived tree canopy height heterogeneity reveals patterns of biodiversity in forest ecosystems. Ecological Informatics, 76, 102082.
